---
title: "Comparing 2D and 3D p-Value Visualizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing 2D and 3D p-Value Visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(p3d)
```

## Introduction

Existing R packages such as `pvaluefunctions` and `concurve` visualize p-values in **two dimensions**. They show how the p-value changes with **one parameter** at a time, typically the effect size, given a fixed standard error.

The **p3d** package extends this to **three dimensions**, showing the p-value as a function of **both** effect size **and** sample size.

This vignette compares the two approaches.

## Setup

We will compare a 2D p-value curve (from `pvaluefunctions`) with a 3D p-value surface (from `p3d`), using the same scenario:

- Observed effect size: *d* = 0.5
- Sample size: *n* = 30

## 2D p-value curve

The `pvaluefunctions` package visualizes the p-value as a function of a single parameter. First, we need to install it:

```{r eval = FALSE}
install.packages("pvaluefunctions")
```

Now we compute the 2D curve:

```{r eval = FALSE}
library(pvaluefunctions)

conf_dist(
  estimate = 0.5,
  stderr = 1 / sqrt(30),
  df = 29,
  tstat = 0.5 * sqrt(30),
  type = "ttest",
  plot_type = "p_val"
)
```

The resulting plot shows the p-value as a function of the estimated effect size, with the observed effect size (0.5) in the center. The curve is **symmetric** and has a **minimum** at the observed effect size.

## 3D p-value surface

The `p3d` package visualizes the p-value as a function of **both** effect size and sample size:

```{r fig.alt = "3D p-value surface"}
surface <- p_surface(
  d = seq(-2, 2, 0.1),
  n = seq(5, 200, 5)
)

plot_p_static(surface)
```

The 3D surface reveals structure that the 2D curve cannot:

1. **The p-value is not symmetric in *n*.** As *n* increases, the surface flattens toward zero for large |*d*| and toward one near *d* = 0.
2. **The significance boundary is a curve, not a point.** For each *n*, there is a critical effect size *d*_crit(*n*) above which the p-value falls below 0.05.
3. **Sample size moderates the effect of *d*.** At small *n*, only large effects are significant. At large *n*, even small effects become significant.

## Comparing the two views

Let's place the 2D curve and the 3D surface side by side. The 2D curve corresponds to a **slice** of the 3D surface at *n* = 30:

```{r fig.alt = "2D curve vs 3D slice"}
# 2D curve at n = 30
curve_30 <- p_curve_2d(d = seq(-2, 2, 0.1), n = 30)

# Plot the 2D curve
plot(curve_30$d, curve_30$p, type = "l", lwd = 2, col = "darkred",
     xlab = "Effect size (d)", ylab = "p-value",
     main = "2D p-value curve (n = 30)")
abline(h = 0.05, col = "blue", lty = 2)
grid()
```

This 2D curve is mathematically equivalent to the **slice** of the 3D surface at *n* = 30. Both give the same p-value at *d* = 0.5:

```{r}
curve_30$p[curve_30$d == 0.5]
surface$p[surface$d == 0.5, surface$n == 30]
```

## What the 2D curve misses

The 2D curve hides the **moderating effect** of sample size. Consider the same effect size (*d* = 0.5) at three different sample sizes:

```{r}
surface$p[surface$d == 0.5, surface$n == 10]
surface$p[surface$d == 0.5, surface$n == 30]
surface$p[surface$d == 0.5, surface$n == 100]
```

At *n* = 10, the p-value is **above** 0.05 (not significant). At *n* = 30, it is **below** 0.05 (significant). At *n* = 100, it is **far below** 0.05 (highly significant).

**This is the moderating effect of sample size.** The same effect size produces different conclusions depending on *n*. The 2D curve cannot show this — it is fixed at one *n*.

## Jensen's inequality: two interpretations of the p-value

The p-value has **two different interpretations**, which the 3D surface helps clarify.

**Conditional p-value** (fixed *d*): the p-value computed for a *specific* effect size. This is what `p_surface()` returns:

$$p_{\text{conditional}}(d, n) = 2 \left[ 1 - F_{t(n-1)}(d \sqrt{n}) \right]$$

For *d* = 0.5, *n* = 30:

```{r}
p_surface(d = 0.5, n = 30)$p[1, 1]
```

**Expected p-value** (random *d*): the average p-value over all possible observed effect sizes:

$$\mathbb{E}[p(d_{\text{obs}}, n)] = \int p(d_{\text{obs}}, n) \cdot f(d_{\text{obs}} \mid d, n) \, dd_{\text{obs}}$$

For *d* = 0.5, *n* = 30, this is approximately:

```{r}
set.seed(42)
n_sim <- 10000
n <- 30
d <- 0.5

p_values <- replicate(n_sim, {
  x <- rnorm(n, mean = d, sd = 1)
  t.test(x, mu = 0)$p.value
})

mean(p_values)
```

**The two values differ substantially** (0.0104 vs 0.0589). This is due to **Jensen's inequality**: the p-value is a convex function of *d*, so the expected p-value is **larger** than the conditional p-value.

The `p3d` package visualizes the **conditional** p-value, consistent with power analysis and existing packages. The **expected** p-value is a subject of ongoing research.

## Conclusion

The 2D and 3D visualizations of the p-value are complementary:

- **2D:** Ideal for examining a single scenario in detail
- **3D:** Ideal for exploring the full parameter space

The 3D surface reveals structure that the 2D curve cannot: the moderating effect of sample size, the continuous nature of the significance boundary, and the interaction between effect size and sample size.

By making these features visible, **p3d** helps weaken the dichotomous thinking that reduces the p-value to a binary decision.

## References

- Helske, J., Helske, S., & Cooper, M. (2021). *Visualizing Hypothesis Tests in Practice*. The R Journal.
- Infanger, D., & Schmidt-Trucksäss, A. (2019). *P value functions: An underused method to present research results and to promote quantitative reasoning*. Statistics in Medicine.
