---
title: "Introduction to p3d"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to p3d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(p3d)
```

## Introduction

The p-value is a continuous measure of statistical evidence, yet in practice it is often reduced to a binary decision: *p* < 0.05 is "significant", *p* ≥ 0.05 is "not significant". This dichotomous thinking contributes to the reproducibility crisis in science.

The **p3d** package visualizes the p-value of a single-sample t-test as a **three-dimensional surface** over two parameters:

- **Effect size** (*d*, Cohen's *d*)
- **Sample size** (*n*)

This vignette demonstrates the core functionality of the package.

## The p-value surface

The central function of the package is `p_surface()`. Given a range of effect sizes and sample sizes, it computes the p-value for each combination:

```{r}
surface <- p_surface(
  d = seq(-2, 2, 0.1),
  n = seq(5, 200, 5)
)

str(surface, max.level = 1)
```

The result is a list with several components:

- `d`: the vector of effect sizes
- `n`: the vector of sample sizes
- `p`: a matrix of p-values (dimensions `length(d)` × `length(n)`)
- `d_crit`: critical effect sizes at α = 0.05 for each *n*
- `alpha`: the significance level
- `grid`: a tidy data frame with all combinations

### Inspecting the results

For example, the p-value at *d* = 0.5, *n* = 30 is:

```{r}
surface$p[surface$d == 0.5, surface$n == 30]
```

The critical effect size at *n* = 30 is:

```{r}
surface$d_crit[surface$n == 30]
```

This means: with 30 observations, an effect size of at least 0.37 is required for statistical significance at α = 0.05.

## Static 3D visualization

The simplest way to visualize the surface is `plot_p_static()`, which uses base R graphics and has **no external dependencies**:

```{r fig.alt = "Static 3D p-value surface"}
plot_p_static(surface)
```

This is ideal for PDF output, journal figures, and printed reports.

## Interactive 3D visualization

For interactive exploration, use `plot_p_surface()` with the `plotly` engine:

```{r eval = FALSE}
plot_p_surface(surface, engine = "plotly")
```

Alternatively, the `rgl` engine provides a native R window:

```{r eval = FALSE}
plot_p_surface(surface, engine = "rgl")
```

Both engines allow rotation, zooming, and panning.

## 2D comparison

For comparison with existing 2D packages such as `pvaluefunctions`, use `p_curve_2d()`:

```{r}
curve_30 <- p_curve_2d(d = seq(-2, 2, 0.1), n = 30)
head(curve_30)
```

The 2D curve is mathematically equivalent to a slice of the 3D surface at fixed *n*.

## Real-data example

Let's apply `p3d` to a real dataset. The `mtcars` dataset contains fuel consumption (mpg) for 32 automobiles. Suppose we want to test whether the mean mpg is different from 20:

```{r}
x <- mtcars$mpg
d_obs <- (mean(x) - 20) / sd(x)
n_obs <- length(x)

cat("Observed effect size:", round(d_obs, 4), "\n")
cat("Sample size:", n_obs, "\n")
```

The p-value for this single observation is:

```{r}
2 * (1 - pt(abs(d_obs) * sqrt(n_obs), df = n_obs - 1))
```

By placing this observation on the 3D surface, we can see how the same effect size would behave at different sample sizes:

```{r fig.alt = "Observed effect size on the 3D surface"}
surface_real <- p_surface(
  d = seq(-0.5, 0.5, 0.01),
  n = seq(5, 200, 5)
)

plot_p_static(surface_real, theta = 30, phi = 20)
```

## Conclusion

The **p3d** package provides a three-dimensional perspective on the p-value. By visualizing the p-value simultaneously across effect size and sample size, it reveals:

1. The **continuous** nature of the p-value (it is not a binary switch)
2. The **moderating effect** of sample size on statistical significance
3. The **interaction** between effect size and sample size

This complements traditional 2D visualizations and helps weaken the dichotomous thinking that plagues statistical practice.

## References

- Helske, J., Helske, S., & Cooper, M. (2021). *Visualizing Hypothesis Tests in Practice*. The R Journal.
- Cohen, J. (1988). *Statistical Power Analysis for the Behavioral Sciences* (2nd ed.). Lawrence Erlbaum Associates.
